Multilevel computation of glacier geometry \\ from Stokes dynamics

### 2021-05-30 paper/msg.tex
Edits section 3.

Though kinematical balance \eqref{eq:ske} applies only on the ice, $\bus \cdot \bn_s + a \le 0$ applies everywhere in $\Omega$ because $a$ is nonpositive in steady state in ice-free locations.  In fact, when combined with the constraint that $s\ge b$ everywhere on $\Omega$, SKE \eqref{eq:ske} is part of an infinite-dimensional nonlinear complementarity problem (NCP) \cite{Bueler2021conservation}.  Recall that an NCP on a finite-dimensional vector space $V=\RR^k$ combines the three statements
\begin{equation}
x\ge 0, \quad F(x)\ge 0, \quad x F(x)=0 \label{eq:ncp}
\end{equation}
for all $x\in V$, where $F:V\to V$ \cite{FacchineiPang2003}.

We may now state the strong form of the SIGP as follows:
\begin{align}
s - b &\ge 0 && \text{on $\Omega$} \label{eq:strongform} \\
- \bu|_s \cdot \bn_s - a &\ge 0 && \text{\emph{same}} \notag \\
(s - b) (- \bu|_s \cdot \bn_s - a) &= 0 && \text{\emph{same}} \notag \\
- \nabla \cdot \left(2 \nu_\eps\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{on $\Lambda_s$} \notag \\
\nabla \cdot \bu &= 0 && \text{\emph{same}} \notag \\
\left(2 \nu_\eps D\bu - pI\right) \bn_s &= \bzero && \text{on $\overline{\partial} \Lambda_s$} \notag \\
\bu &= \bzero && \text{on $\underline{\partial} \Lambda_s$} \notag
\end{align}
(Definition \eqref{eq:regeffvisc} is also needed.)  The first three statements in \eqref{eq:strongform} form the NCP, but it is coupled to the dynamical boundary value problem formed by the last four statements.

The solution of \eqref{eq:strongform} is a triple of functions $s(x,y)$, $\bu(x,y,z)$, $p(x,y,z)$.  However, as the domain on which $\bu,p$ are defined is only known via the solution elevation $s$, \eqref{eq:strongform} is at best an incomplete description.  The weak form in the next section, using a $s\mapsto \bu|_s$ solution operator, will partially address this concern.  Whether in strong or weak form, this inequality-constrained system of PDEs has a largely-unknown theory regarding well-posedness and solution regularity.  However, certain well-posedness theory is known for the SIA version of this problem, with existence established by \cite{JouvetBueler2012}, and uniqueness in the flat bed case.

In Sections \ref{sec:fe}--\ref{sec:mcdstokes} we will construct and demonstrate a robust multilevel scheme for the numerical solution of the SIGP weak form (Section \ref{sec:weakido}).  As noted in Section \ref{sec:intro}, numerical solutions have traditionally applied explicit time-stepping, even when computing steady states, splitting the dynamics and the SKE into sub-steps and using truncation to address the NCP \cite[for example]{Jouvetetal2008,Lengetal2012}.  By contrast, the implicit time-stepping model in Section \ref{sec:evolution} avoids time-splitting and is unconditionally stable.

\section{Weak form and the ice dynamics operator} \label{sec:weakido}

The weak form of the fixed-geometry Glen-Stokes model for ice flow, which extends the linear Stokes weak form \cite{Elmanetal2014} to regularized power-law rheology, is relatively well-known \cite{IsaacStadlerGhattas2015,JouvetRappaz2011,Lengetal2012}, and we summarize it here.  Then we will address the weak form of the SIGP.

Denote by $W^{\text{k},\text{r}}$ the Sobolev space \cite{Evans2010} of functions with $k$ derivatives which are $r$th-power integrable.  Suppose for the moment that $\Lambda \subset \RR^{d+1}$ is a fixed, bounded domain on which we prescribe a Dirchlet condition $\bu=0$ on some positive-measure portion of the boundary, and otherwise suppose the boundary is stress-free (Neumann boundary).  Let $\pp=(1/n)+1$ as in \eqref{eq:regeffvisc}, and let $\qq=(1-\pp^{-1})^{-1}=n+1$ be the conjugate exponent; $\pp=4/3$ and $\qq = 4$ if $n=3$.  Let $W_0^{1,\pp}(\Lambda)^{d+1}$ be the space of velocity functions, zero along the Dirichlet boundary, and
\begin{equation}
\mathcal{M}_{\text{GS}} = W_0^{1,\pp}(\Lambda)^{d+1} \times L^\qq(\Lambda)  \label{eq:mixed}
\end{equation}
be the (mixed) space of admissible velocity and pressure pairs.  The Glen-Stokes weak form is the statement that $(\bu,p) \in \mathcal{M}_{\text{GS}}$ should satisfy
\begin{equation}
F_{\text{GS}}(\bu,p)[\bv,q] = \int_\Lambda 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0 \label{eq:glenstokesweak}
\end{equation}
for all $(\bv,q) \in \mathcal{M}_{\text{GS}}$.

Jouvet and Rappaz \cite{JouvetRappaz2011} prove that this fixed-domain Glen-Stokes formulation is well-posed under the above assumptions if also the Neumann boundary is $C^1$.  In particular they show \eqref{eq:glenstokesweak} is equivalent to minimization of a convex and coercive functional over the divergence-free subspace, but also that there is a unique pressure $p\in L^\qq(\Lambda)$.

However, we seek to go beyond this dynamical problem on predetermined ice geometry.  We are interested in finding a triple $s,\bu,p$ which solves \eqref{eq:strongform}, or rather its weak form which we now propose.  Again suppose $\Omega \subset \RR^d$ is a fixed map-plane domain with $C^1$ boundary, and that the bed elevation $b$ is in $W^{1,\qq}(\Omega)$.  Because of the $C^1$ boundary, the Sobolev space $W^{1,\qq}(\Omega)$ supports a linear trace operator which defines $f|_{\partial \Omega}$ for $f \in W^{1,\qq}(\Omega)$ \cite[Section 5.5]{Evans2010}.  Let
\begin{equation}
\mathcal{K} = \{s \in W^{1,\qq}(\Omega) \,:\, s \ge b \, \text{ and } \, s\big|_{\partial\Omega} = b\big|_{\partial\Omega}\}  \label{eq:Kconstraintset}
\end{equation}
be the closed and convex set of admissible surface elevations of an isolated glacier or ice sheet.  However, it is not clear that the correct Sobolev space has been identified in \eqref{eq:Kconstraintset}, and in fact we must further suppose that
\begin{equation}
s\in \mathcal{K} \text{ defines an icy domain } \Lambda_s \text{ which has a $C^1$ free surface.} \label{eq:quixotic}
\end{equation}
Quixotic assumption \eqref{eq:quixotic} would follow from a sufficiently-strong regularity result for the theory of the current paper, which we cannot offer.  However, the existing theory of shallow ice sheet well-posedness \cite{JouvetBueler2012} includes a version of definition \eqref{eq:Kconstraintset},\footnote{In \cite{JouvetBueler2012} the $2\qq/(\qq-1)=8/3$ power of the ice thickness is in $W^{1,\qq}(\Omega)$.} and in that shallow theory $s$ solves an elliptic equation and is thus smooth, at least interior to the free boundary.

We may now define the ice dynamics operator (IDO), a map using the trace of the velocity solution to the Glen-Stokes problem above (Figure \ref{fig:idoaction}).  For each surface elevation function $s \in \mathcal{K}$ the IDO result $\Phi(s)$ is the normal component of the upper-surface trace of the unique velocity solution $\bu$ of \eqref{eq:glenstokesweak} over the icy domain $\Lambda_s$, but extended by zero to all of $\Omega$:
\begin{equation}
\Phi(s) = \begin{cases} \bu|_{\overline{\partial} \Lambda_s} \cdot \bn_s, & s > b, \\
                        0, & \text{otherwise}, \end{cases} \label{eq:ido}
\end{equation}
where $\bu|_{\overline{\partial} \Lambda_s}$ denotes a trace operator on $W^{1,\pp}(\Lambda_s)^{d+1}$; compare \eqref{eq:mixed}.  Note that the output $\Phi(s)$, a scalar function on $\Omega$, may jump or have other singularities at the glacier margin, but we will only need $\Phi(s)$ to be in the dual of $W^{1,\qq}(\Omega)$, thus permitting substantial irregularities.

\begin{figure}[t]
\begin{center}
\includegraphics[width=\textwidth]{genfigs/idoaction.pdf}
\end{center}
\caption{The ice dynamics operator $\Phi$ maps the surface elevation $s$ (dashed) to the normal ice surface motion $\Phi(s)=\bu|_s \cdot \bn_s$; the thin line is $s+\Phi(s)$.  Both the input and output of $\Phi$ are defined on all of $\Omega$.}
\label{fig:idoaction}
\end{figure}

@@

For $s \in \mathcal{K}$ and $r \in W^{1,\qq}(\Omega)$ we next define a functional which is nonlinear in $s$:
\begin{equation}
F(s)[r] = - \ip{\Phi(s)}{r} = - \int_\Omega \Phi(s)\, r \,dx dy. \label{eq:sigpfunctional}
\end{equation}
One computes $F(s)[r]$ by defining the icy domain $\Lambda_s$ \eqref{eq:lambdas}, then solving the (weak form) Glen-Stokes problem \eqref{eq:glenstokesweak}, then evaluating the normal component of the surface trace of the velocity \eqref{eq:ido} to compute the IDO $\Phi(s)$, and then integrating the result against the test function $r$.  Implicit in \eqref{eq:sigpfunctional} is that $\Phi(s)$ has minimal regularity; it need only be in the dual space of $W^{1,\qq}(\Omega)$.

Finally, the SIGP weak form is the following variational inequality (VI) \cite{KinderlehrerStampacchia1980} which should determine $s\in\mathcal{K}$:
\begin{equation}
F(s)[r - s] \ge \ip{a}{r-s} \quad \text{for all $r \in \mathcal{K}$.}  \label{eq:sigpweakform}
\end{equation}
This VI has a shallow (SIA) analog, for which existence has been proven \cite{JouvetBueler2012}, but in that case the functional is an integral of powers of the thickness $s-b$ and the surface slope $\grad s$, thus locally-computable.  Here, by contrast, the computation of \eqref{eq:sigpfunctional} is non-local and expensive: to evaluate $F(s)[r]$ one must solve Glen-Stokes problem \eqref{eq:glenstokesweak} on the given $z=s(x,y)$ geometry.

Definition \eqref{eq:sigpfunctional} is a dual pairing of $F(s)$ and $r$, thus roughly an inner product.  VI \eqref{eq:sigpweakform} says that $s$ is located in $\mathcal{K}$, generically on $\partial\mathcal{K}$, at a place where $F(s)$ points directly into $\mathcal{K}$.  Specifically, in geometrical analogy, \eqref{eq:sigpweakform} says that the ``angle'' between $F(s)$ and an arbitrary vector $r-s$ pointing into $\mathcal{K}$ is at most $90^\circ$.

It is well-known that some VIs arise as inequality-constrained minimization problems \cite{GraeserKornhuber2009,KinderlehrerStampacchia1980}, but to the best of our knowledge VI \eqref{eq:sigpweakform} does \emph{not} arise in this way.  The general-bed, steady shallow (SIA) model is not a minimization, except in the flat-bed case \cite{JouvetBueler2012}.

The SIGP weak form \eqref{eq:sigpweakform} is coupled to \eqref{eq:glenstokesweak} through the definition of the IDO $\Phi$.  The problem therefore has three fundamental nonlinearities:
\renewcommand{\labelenumi}{(\emph{\roman{enumi}})}
\begin{enumerate}
\item the Glen power-law rheology,
\item the inequality constraint, and
\item the nonlinearity of free-surface flow reflected in the nonlinearity of the operator $\Phi$.
\end{enumerate}
Regarding (\emph{ii}), observe that even the classical obstacle problem for the linear Laplacian operator is nonlinear \cite{KinderlehrerStampacchia1980}.  Regarding (\emph{iii}), note that a free-surface flow for Newtonian (linear) rheology produces a nonlinear equation for the flow thickness, e.g.~as expressed in the kinematic wave equation \cite{Ockendonetal2003}.

To numerically solve \eqref{eq:sigpweakform} we will need the (Gateaux) derivative of $F$, which requires differentiating the $\Phi$ as well.  Suppose $s\in \mathcal{K}$, $\eps>0$, and $t \in W^{1,\qq}(\Omega)$ is such that $s+\eps t \in \mathcal{K}$.  (Both $s$ and $s+\eps t$ must be admissible glacier surface elevations.)  From \eqref{eq:sigpfunctional},
\begin{equation}
F'(s)[t,r] = \lim_{\eps\to 0^+} \frac{F(s+\eps t)[r] - F(s)[r]}{\eps} = - \int_\Omega \Phi'(s)[t]\, r \,dx dy \label{eq:sigpfunctionalderiv}
\end{equation}
where by definition
\begin{equation}
\Phi'(s)[t] = \lim_{\eps\to 0^+} \frac{\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s}{\eps}. \label{eq:idoderiv}
\end{equation}
In \eqref{eq:idoderiv} note that the normal vectors and surface velocities are extended to the whole of $\Omega$; see \eqref{eq:surfacevelocity}.  We will assume that $\Phi'(s)[t]$ is in the dual space of $W^{1,\qq}(\Omega)$ so that \eqref{eq:sigpfunctionalderiv} is finite.  Observe that $F'$ is straightforward to compute from $\Phi'$.

Note $F'(s)[t,r]$ and $\Phi'(s)[t]$ are one-sided directional derivatives in the sense that $s+\eps t$ must be admissible, i.e.~$s+\eps t\ge b$, for all sufficiently-small $\eps>0$, and this clearly requires $t\ge 0$ on the (active) set where $s=b$.  However, to simplify our considerations we define an $s$-independent set
\begin{equation}
\mathcal{D}_+ = \{t \in W^{1,\qq}(\Omega) \,:\, t(x,y) \ge 0\}, \label{eq:infdefectset}
\end{equation}
so $\Phi'(s)[t]$ is well-defined on inputs $(s,t) \in \mathcal{K} \times \mathcal{D}_+$ and $F'(s)[t,r]$ is well-defined for $(s,t,r) \in \mathcal{K} \times \mathcal{D}_+ \times W^{1,\qq}(\Omega)$.

We will approximate the true derivative $\Phi'(s)[t]$ by the difference quotient in \eqref{eq:idoderiv}, but only in the situation where $t \in \mathcal{D}_+$ has small support.  (It will be a finite element basis function; see the next section.)  For this it is relevant to say more about the difference, as it depends on whether the point is in the support of $t$ or not:
\begin{align}
\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s &= \left(\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_{s+\eps t}\right) + \left(\bus \cdot \bn_{s+\eps t} - \bus \cdot \bn_s\right) \label{eq:differencecases} \\
    &= \begin{cases}
           (\bu|_{s+\eps t} - \bus) \cdot \bn_{s+\eps t} + \eps\, \bus \cdot \left<t_x,t_y,0\right>, & s > b \\
           \bu|_{s+\eps t} \cdot \bn_{s+\eps t}, & s=b \text{ and } t > 0 \\
           0, & s=b \text{ and } t = 0
                 \end{cases} \notag
\end{align}
Thus a necessary condition for differentiability of $\Phi$ is that $\bu|_{s+\eps t} \cdot \bn_{s+\eps t} = o(\eps)$ on the ice-free area $s=b$.  In physical terms this means that the dynamical part of the SKE vanishes for small ice masses on bare ground, and it is reasonable to suppose this is so if the ice cannot slide.

In the next three sections we will construct an iterative, multilevel finite element solver for SIGP weak form \eqref{eq:sigpweakform}.  When we approximate the true derivative $\Phi'(s)[t]$ by a difference quotient, a significant issue arises because two Glen-Stokes solutions over the ice domains $\Lambda_{s+\eps t}$ and $\Lambda_s$ are seemingly required.  While this would imply a very large numbers of expensive dynamical solves, in Section \ref{sec:smoothers} we propose a less-expensive approximation.

>>>>

Though kinematical balance \eqref{eq:ske} applies only on the ice, $\bus \cdot \bn_s + a \le 0$ applies everywhere in $\Omega$ because $a$ is nonpositive in steady state in ice-free locations.  In fact, when combined with the constraint that $s\ge b$ everywhere on $\Omega$, SKE \eqref{eq:ske} is part of an infinite-dimensional nonlinear complementarity problem (NCP) \cite{Bueler2021conservation}.  An NCP on a finite-dimensional vector space $V=\RR^k$ combines the three statements
\begin{equation}
x\ge 0, \quad f(x)\ge 0, \quad x f(x)=0 \label{eq:ncp}
\end{equation}
for all $x\in V$, where $f:V\to V$ \cite{FacchineiPang2003}.

We may now state the strong form of the SIGP, as follows, by including all of the above conditions, and also by eliminating $\tau$ from \eqref{eq:forcebalance} and \eqref{eq:viscflowlaw}:
\begin{align}
s - b &\ge 0 && \text{on $\Omega$} \label{eq:strongform} \\
- \bu|_s \cdot \bn_s - a &\ge 0 && \text{\emph{same}} \notag \\
(s - b) (- \bu|_s \cdot \bn_s - a) &= 0 && \text{\emph{same}} \notag \\
- \nabla \cdot \left(2 \nu_\eps\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{on $\Lambda_s$} \notag \\
\nabla \cdot \bu &= 0 && \text{\emph{same}} \notag \\
\left(2 \nu_\eps D\bu - pI\right) \bn &= \bzero && \text{on $\partial \Lambda_s \setminus \Gamma_0$} \notag \\
\bu &= \bzero && \text{on $\Gamma_0$} \notag
\end{align}
(Definition \eqref{eq:regeffvisc} of $\nu_\eps$ is also needed.)  The first three statements in \eqref{eq:strongform} form the NCP, but this problem is coupled to the boundary value problem formed by the last four statements.

The solution of \eqref{eq:strongform} is a triple of functions $s(x,y)$, $\bu(x,y,z)$, $p(x,y,z)$ on $\Omega,\Lambda_s,\Lambda_s$, respectively.  However, as the domain on which $\bu,p$ are defined is only known via the solution, \eqref{eq:strongform} is at best an incomplete description.  The weak form in the next section, using a solution operator which maps $s$ to $\bu|_s$, will partially address this concern.

However, whether in strong or weak form, inequality-constrained system \eqref{eq:strongform} has a largely-unknown theory regarding well-posedness and solution regularity.  Certain well-posedness theory is known for the SIA version of this problem, with existence established by \cite{JouvetBueler2012}, including uniqueness in the flat bed case.

After stating the SIGP weak form next in Section \ref{sec:weakido}, in Sections \ref{sec:fe}--\ref{sec:results} we will construct and demonstrate a multilevel scheme for the problem.  As noted in Section \ref{sec:intro}, numerical solutions have traditionally applied explicit time-stepping, even when computing steady states, splitting the dynamics and the SKE into sub-steps and using truncation to address the NCP \cite[for example]{Jouvetetal2008,Lengetal2012}.  By contrast, the implicit time-stepping model in Section \ref{sec:evolution} avoids time-splitting and is unconditionally stable.

\section{The weak form using an ice dynamics operator} \label{sec:weakido}

The weak form of the fixed-geometry Glen-Stokes model for ice flow, which extends the linear Stokes weak form \cite{Elmanetal2014} to regularized power-law rheology, is relatively well-known \cite{IsaacStadlerGhattas2015,JouvetRappaz2011,Lengetal2012}, and we summarize it here, after which we address the weak form of the SIGP.

Denote by $W^{k,r}$ the Sobolev space \cite{Evans2010} of functions with $k$ derivatives which are $r$th-power integrable.  Suppose for the moment that $\Lambda \subset \RR^{d+1}$ is a fixed, bounded domain on which we prescribe a Dirchlet condition $\bu=0$ on some positive-measure portion of the boundary, and otherwise suppose the boundary is stress-free (Neumann boundary).  Let $\pp=(1/n)+1$ as in \eqref{eq:regeffvisc}, and $\qq=(1-\pp^{-1})^{-1}=n+1$ the conjugate exponent; $\pp=4/3$ and $\qq = 4$ if $n=3$.  Let $W_0^{1,\pp}(\Lambda)^{d+1}$ be the space of velocity functions, zero along the Dirichlet boundary, and let
\begin{equation}
\mathcal{M}_{\text{GS}} = W_0^{1,\pp}(\Lambda)^{d+1} \times L^\qq(\Lambda)  \label{eq:mixed}
\end{equation}
be the (mixed) space of admissible velocity and pressure pairs.  The Glen-Stokes solution $(\bu,p) \in \mathcal{M}_{\text{GS}}$ should satisfy the weak form
\begin{equation}
F_{\text{GS}}(\bu,p)[\bv,q] = \int_\Lambda 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0 \label{eq:glenstokesweak}
\end{equation}
for all $(\bv,q) \in \mathcal{M}_{\text{GS}}$.

Jouvet and Rappaz \cite{JouvetRappaz2011} prove that this fixed-domain Glen-Stokes formulation is well-posed under the above assumptions if also the Neumann boundary of $\Lambda$ is $C^1$.  In particular they show \eqref{eq:glenstokesweak} is equivalent to minimization of a convex and coercive functional over the divergence-free subspace, but also that there is a unique pressure $p$.

However, we must go beyond the problem on predetermined ice geometry to solve the weak form of \eqref{eq:strongform}.  Again suppose $\Omega \subset \RR^d$ is a fixed map-plane domain with $C^1$ boundary, and that the bed elevation $b$ is in $W^{1,\qq}(\Omega)$.  The Sobolev space $W^{1,\qq}(\Omega)$ supports a linear trace operator which defines the symbol $|_{\partial \Omega}$ \cite[Section 5.5]{Evans2010}.  Let
\begin{equation}
\mathcal{K} = \{s \in W^{1,\qq}(\Omega) \,:\, s \ge b \, \text{ and } \, s\big|_{\partial\Omega} = b\big|_{\partial\Omega}\}  \label{eq:Kconstraintset}
\end{equation}
be the closed and convex set of admissible surface elevations of an isolated glacier or ice sheet.  From now on we also assume that $n>1$.  Because $\qq = n+1 > d$, each $s \in W^{1,\qq}(\Omega)$ is continuous \cite[Morrey's inequality, section 5.6.2]{Evans2010}, and so there are no cliffs in the $s$ geometry.

However, it is not clear that the correct Sobolev space has been identified in \eqref{eq:Kconstraintset}, and in fact we must further suppose that
\begin{equation}
s\in \mathcal{K} \text{ defines an icy domain } \Lambda_s \text{ which has a $C^1$ upper surface.} \label{eq:quixotic}
\end{equation}
Quixotic assumption \eqref{eq:quixotic} would follow from a sufficiently-strong regularity result for the theory of the current paper, which we cannot offer.  However, the existing theory of well-posedness for the SIA model includes a version of definition \eqref{eq:Kconstraintset} \cite{JouvetBueler2012}, wherein a power of the ice thickness, $u=H^{2q/(q-1)}$, is in $\mathcal{K} = \{u \ge 0\} \subset W^{1,\qq}(\Omega)$, and furthermore in the SIA the surface $s$ solves an elliptic equation and is thus smooth on the set $\{s<b\} \subset \Omega$.

We may now define the ice dynamics operator (IDO), a map which uses the trace of the velocity solution to the Glen-Stokes problem above (Figure \ref{fig:idoaction}), as follows.  For each surface elevation function $s \in \mathcal{K}$ the IDO result $\Phi(s)$ is the normal component of the upper-surface trace of the unique velocity solution $\bu$ of \eqref{eq:glenstokesweak} over the icy domain $\Lambda_s$, but extended by zero to all of $\Omega$:
\begin{equation}
\Phi(s) = \begin{cases} \bu|_{\overline{\partial} \Lambda_s} \cdot \bn_s, & s > b, \\
                        0, & \text{otherwise}. \end{cases} \label{eq:ido}
\end{equation}
Here $\bu|_{\overline{\partial} \Lambda_s}$ denotes a trace operator on $W^{1,\pp}(\Lambda_s)^{d+1}$; compare \eqref{eq:mixed}.  Note that the output $\Phi(s)$, a scalar function on $\Omega$, may jump or have other singularities at the glacier margin, but we will only need $\Phi(s)$ to be in the dual of $W^{1,\qq}(\Omega)$, permitting substantial irregularities.

\begin{figure}[t]
\begin{center}
\includegraphics[width=\textwidth]{genfigs/idoaction.pdf}
\end{center}
\caption{The ice dynamics operator $\Phi$ maps the surface elevation $s$ (dashed) to the normal ice surface motion $\Phi(s)=\bu|_s \cdot \bn_s$, equation \eqref{eq:ido}.  Both the input and output of $\Phi$ are defined on all of $\Omega$.}
\label{fig:idoaction}
\end{figure}

@@

For $s \in \mathcal{K}$ and $r \in W^{1,\qq}(\Omega)$ we next define a functional which is nonlinear in $s$:
\begin{equation}
F(s)[r] = - \ip{\Phi(s)}{r} = - \int_\Omega \Phi(s)\, r \,dx dy. \label{eq:sigpfunctional}
\end{equation}
One computes $F(s)[r]$ by defining the icy domain $\Lambda_s$ as in \eqref{eq:lambdas}, then solving the (weak form) Glen-Stokes problem \eqref{eq:glenstokesweak}, then evaluating the normal component of the surface trace of the velocity \eqref{eq:ido} to compute the IDO $\Phi(s)$, and then integrating the result against the test function $r$.  Implicit in \eqref{eq:sigpfunctional} is that $\Phi(s)$ is in the dual space of $W^{1,\qq}(\Omega)$, so it possesses minimal regularity.

Finally, the SIGP weak form is the following variational inequality (VI) \cite{KinderlehrerStampacchia1980} which should determine $s\in\mathcal{K}$:
\begin{equation}
F(s)[r - s] \ge \ip{a}{r-s} \quad \text{for all $r \in \mathcal{K}$.}  \label{eq:sigpweakform}
\end{equation}
This VI has a shallow (SIA) analog, for which existence has been proven \cite{JouvetBueler2012}, but in that case the functional is an integral of powers of the thickness $s-b$ and the surface slope $\grad s$, thus locally-computable.  Here, by contrast, the computation of \eqref{eq:sigpfunctional} is non-local and expensive: to evaluate $F(s)[r]$ one must solve Glen-Stokes problem \eqref{eq:glenstokesweak} on the given $z=s(x,y)$ geometry.  In any case one can prove that if $s$ is $C^1$ and solves \eqref{eq:sigpweakform} then it solves strong forms \eqref{eq:strongform} and \eqref{eq:idostrongform} \cite{Bueler2021conservation}.  In particular the strong SKE \eqref{eq:ske} holds on $s>b$.

Definition \eqref{eq:sigpfunctional} is a dual pairing of $F(s)$ and $r$, thus roughly an inner product.  Speaking geometrically, VI \eqref{eq:sigpweakform} says that $s$ is located in $\mathcal{K}$, generically on $\partial\mathcal{K}$, at a place where $F(s)$ points directly into $\mathcal{K}$.  Specifically, \eqref{eq:sigpweakform} says that the ``angle'' between $F(s)$ and an arbitrary vector $r-s$ pointing into $\mathcal{K}$ is at most $90^\circ$.

It is well-known that certain VIs arise as inequality-constrained minimization problems \cite{GraeserKornhuber2009,KinderlehrerStampacchia1980}, but to the best of our knowledge VI \eqref{eq:sigpweakform} does \emph{not} arise in this way.  The general-bed, steady shallow (SIA) model is also not a minimization, except in the flat-bed case \cite{JouvetBueler2012}.

The SIGP weak form \eqref{eq:sigpweakform} is coupled to \eqref{eq:glenstokesweak} through definition \eqref{eq:ido} of the IDO $\Phi$.  The problem therefore has three fundamental nonlinearities:
\renewcommand{\labelenumi}{(\emph{\roman{enumi}})}
\begin{enumerate}
\item the Glen power-law rheology,
\item the inequality constraint, and
\item the nonlinearity of free-surface flow reflected in the nonlinearity of the operator $\Phi$.
\end{enumerate}
Regarding (\emph{ii}), observe that even the classical obstacle problem for the linear Laplacian operator is nonlinear \cite{KinderlehrerStampacchia1980}.  Regarding (\emph{iii}), note that a free-surface flow for Newtonian (linear) rheology produces a nonlinear equation for the flow thickness, e.g.~as expressed in the kinematic wave equation \cite{Ockendonetal2003}.

We will need the (Gateaux) derivative of $F$, and this requires differentiating $\Phi$ as well.  Suppose $s\in \mathcal{K}$, $\eps>0$, and $t \in W_0^{1,\qq}(\Omega)$ is such that $s+\eps t \in \mathcal{K}$.  (In the following, both $s$ and $s+\eps t$ must be admissible glacier surface elevations.)  From \eqref{eq:sigpfunctional}, if $r\in W_{1,\qq}(\Omega)$ then
\begin{equation}
F'(s)[t,r] = \lim_{\eps\to 0^+} \frac{F(s+\eps t)[r] - F(s)[r]}{\eps} = - \int_\Omega \Phi'(s)[t]\, r \,dx dy \label{eq:sigpfunctionalderiv}
\end{equation}
where by definition
\begin{equation}
\Phi'(s)[t] = \lim_{\eps\to 0^+} \frac{\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s}{\eps}. \label{eq:idoderiv}
\end{equation}
In \eqref{eq:idoderiv} note that all quantities are extended to the whole of $\Omega$.

We will assume that $\Phi'(s)[t]$ is in the dual space of $W^{1,\qq}(\Omega)$ so that \eqref{eq:sigpfunctionalderiv} is finite.  Observe that $F'$ is straightforward to compute from $\Phi'$, but further comments on \eqref{eq:idoderiv} are forthcoming.  In fact, $F'(s)[t,r]$ and $\Phi'(s)[t]$ are one-sided directional derivatives in the sense that $s+\eps t$ must be admissible, i.e.~$s+\eps t\ge b$, for all sufficiently-small $\eps>0$, which requires $t\ge 0$ on the (active) set where $s=b$.  However, to simplify our considerations we define an $s$-independent set
\begin{equation}
\mathcal{D}_+ = \{t \in W_0^{1,\qq}(\Omega) \,:\, t(x,y) \ge 0\}. \label{eq:infdefectset}
\end{equation}
It follows that $\Phi'(s)[t]$ is well-defined on inputs $(s,t) \in \mathcal{K} \times \mathcal{D}_+$ and $F'(s)[t,r]$ is well-defined for $(s,t,r) \in \mathcal{K} \times \mathcal{D}_+ \times W^{1,\qq}(\Omega)$.  We will approximate the true derivative $\Phi'(s)[t]$ by the difference quotient in \eqref{eq:idoderiv}, though only in the situation where $t \in \mathcal{D}_+$ has small support.

In fact the meaning of the finite difference in \eqref{eq:idoderiv} depends on the support of $t$:
\begin{equation}
\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s = \begin{cases}
           (\bu|_{s+\eps t} - \bus) \cdot \bn_{s+\eps t} + \eps\, \bus \cdot \left<t_x,t_y,0\right>, & s > b \\
           \bu|_{s+\eps t} \cdot \bn_{s+\eps t}, & s=b, t > 0 \\
           0, & s=b, t = 0.
                 \end{cases} \label{eq:differencecases}
\end{equation}
Thus a necessary condition for differentiability of $\Phi$ is that $\bu|_{s+\eps t} \cdot \bn_{s+\eps t} = o(\eps)$ on the ice-free area $s=b$.  In physical terms this means that the dynamical part of the SKE \eqref{eq:ske} vanishes for small ice masses on bare ground, which is reasonable to suppose if the ice cannot slide.

In the next three sections we will construct an iterative, multilevel finite element solver for SIGP weak form \eqref{eq:sigpweakform}.  When we attempt to approximate the derivative $\Phi'(s)[t]$ by a difference quotient, a significant issue arises because an extra solution of Glen-Stokes problem \eqref{eq:glenstokesweak} is required over the icy domain $\Lambda_{s+\eps t}$, but in Section \ref{sec:smoothers} we propose a less-expensive approximation.
